---
title: "Degree Day Model"
author: "Caitlynn Tran & Marianna Karageorgi"
date: "01/10/2023"
---

## Background

The purpose of the present code is to use the Degree-Day Model to estimate and track number of generations for the field experiment in 2021. We use the model parameters as in Behrman & Schmidt 2022.

Behrman estimated that *D. melanogaster* required approximately 112 degree days to develop from egg to adult.

**The model:**

Average model = (Daily max temp + Daily min temp)/2 - baseline

-   When tmin \< tbase, replace with tbase

-   When tmin \< tbase, replace with tbase & when tmax \> tbase_max, replace by tbase_max

-   When tmin \< tbase, replace with tbase & when tmax \> tbase_max, no units added

```{r}
# load packages for data wrangling 
library(dplyr)
library(Rmisc)
library(tidyverse)
library(scales)
```

## 1. Generations in field mesocosms 

```{r}
start_date = as.Date("2021-06-21")

weather = read.csv("./data/2021_orchard_weather_data.csv")%>%
  select("TIMESTAMP", "Temp_C_Avg", "Temp_C")%>%
  #Since the date is formatted like "Thursday, April 1, 2021" we want to reformat it in a way so the code can understand it as a date. We first separate the value using the delimiter "," and then separate the month and day using a space as a delimiter
  separate("TIMESTAMP", c("Day of the Week", "Month + Day", "Year"), ", ")%>%
  separate("Month + Day", c("Month", "Day"), " ")%>%
  select("Month":"Temp_C")%>% #only select for Month, Day, Year, and temperature variables (drop the day of the week)
  
# We want to replace all the months with their number equivalent (so April is the fourth month of the year, so we replace April with 4)
  mutate(Month = str_replace(Month, "April", "4"))%>%
  mutate(Month = str_replace(Month, "May", "5"))%>%
  mutate(Month = str_replace(Month, "June", "6"))%>%
  mutate(Month = str_replace(Month, "July", "7"))%>%
  mutate(Month = str_replace(Month, "August", "8"))%>%
  mutate(Month = str_replace(Month, "September", "9"))%>%
  mutate(Month = str_replace(Month, "October", "10"))%>%
  mutate(Month = str_replace(Month, "November", "11"))%>%
  mutate(Month = str_replace(Month, "December", "12"))%>%
  drop_na()%>% # This gets rid of some extra rows at the top of the dataset
 
  
# This combines the Month, Day and Year together into one column, separating them with a slash so it looks like 4/1/2021. Then we use the as.Date() function to store the value as a date
  unite("Date", Month:Year, sep = "/")%>%
  mutate(Date = as.Date(Date, "%m/%d/%Y"))%>% 
  filter(Date - start_date >= 0)%>%
  mutate(Temp_C_Avg = as.double(Temp_C_Avg), Temp_C = as.double(Temp_C))%>% #want to store these values as number values not as characters
  
  # We group each of the values by the day so we are able to calculate the average temp for each day
  dplyr::group_by(Date)%>%
  dplyr::summarize(Low = min(Temp_C_Avg), High = max(Temp_C_Avg))


weather
```

```{r}
# Set temperature thresholds and baseline for degree day calculation
baseline = 12
max_high = 29
min_low = 12
threshold = 112

# Function to calculate the average degrees per day
ddm <- function(low, high) {
  average_model <- (low + high)/2 - baseline
  return(average_model)
}

# Create a new column assigning the value as calculated from the average model. If the High is greater than the max_high, 29 will be used to calculate. If the Low is less than the min_low, 12 will be used to calculate
avg_model <- weather %>%
  mutate("Average_Model" = case_when(
    (Low < min_low & High < min_low) ~ ddm(min_low, min_low),
    (Low >= max_high & High >= max_high) ~ ddm(max_high, max_high),
    (Low >= min_low & High <= max_high) ~ ddm(Low, High), 
    (Low < min_low & High <= max_high) ~ ddm(min_low, High),
    (Low >= min_low & High > max_high) ~ ddm(Low, max_high),
    (Low < min_low & High > max_high) ~ ddm(min_low, max_high),
    TRUE ~ ddm(Low, High)
  ))

# Variable to keep track of the cumulative average
cumulative_average = 0

# Empty vector for the generation dates to be added to
generation_dates = c()

# The for loop calculates the cumulative average 
for (num in 1:length(avg_model$Average_Model)) {
  # this takes the average model value calculated above and adds to the cumulative average variable
  cumulative_average = cumulative_average + avg_model$Average_Model[num]
 
  # if the cumulative average exceeds the threshold of 112, it will take that date and add it to the generation_dates vector. it also resets the cumulative average
   if (cumulative_average >= threshold) {
    generation_dates = append(generation_dates, as.Date(avg_model$Date[num]))
    cumulative_average = 0
  }
}

print(generation_dates)

# Create new column with generation information and pesticide information
generation_start_June21 = avg_model %>%
  mutate(generation = case_when(
    Date >= as.Date("2021-06-21") & Date < as.Date("2021-06-30") ~ 1,
    Date >= as.Date("2021-06-30") & Date < as.Date("2021-07-10") ~ 2,
    Date >= as.Date("2021-07-10") & Date < as.Date("2021-07-19") ~ 3,
    Date >= as.Date("2021-07-19") & Date < as.Date("2021-07-29") ~ 4,
    Date >= as.Date("2021-07-29") & Date < as.Date("2021-08-09") ~ 5,
    Date >= as.Date("2021-08-09") & Date < as.Date("2021-08-18") ~ 6,
    Date >= as.Date("2021-08-18") & Date < as.Date("2021-08-27") ~ 7,
    Date >= as.Date("2021-08-27") & Date < as.Date("2021-09-07") ~ 8,
    Date >= as.Date("2021-09-07") & Date < as.Date("2021-09-18") ~ 9,
    Date >= as.Date("2021-09-18") & Date < as.Date("2021-10-01") ~ 10,
    Date >= as.Date("2021-10-01") & Date < as.Date("2021-10-14") ~ 11,
    Date >= as.Date("2021-10-14") & Date < as.Date("2021-11-09") ~ 12,

    TRUE ~ NA_integer_  # Assign NA for dates outside the specified ranges
    ),
      dose_ppm = case_when(
    Date >= as.Date("2021-06-21") & Date <= as.Date("2021-07-20") ~ 0,
    Date >= as.Date("2021-07-21") & Date <= as.Date("2021-08-24") ~ 2.5,
    Date >= as.Date("2021-08-25") & Date <= as.Date("2021-09-06") ~ 7.5,
    Date >= as.Date("2021-09-07") & Date <= as.Date("2021-12-09") ~ 0,
    TRUE ~ NA_real_  # Assign NA for dates outside the specified ranges
    ))


write.csv(generation_start_June21,"./data/generation_start_June21.csv")
```

## 2. Generations for simulations

```{r}
start_date = as.Date("2021-07-13")

weather = read.csv("./data/2021_orchard_weather_data.csv")%>%
  select("TIMESTAMP", "Temp_C_Avg", "Temp_C")%>%
  #Since the date is formatted like "Thursday, April 1, 2021" we want to reformat it in a way so the code can understand it as a date. We first separate the value using the delimiter "," and then separate the month and day using a space as a delimiter
  separate("TIMESTAMP", c("Day of the Week", "Month + Day", "Year"), ", ")%>%
  separate("Month + Day", c("Month", "Day"), " ")%>%
  select("Month":"Temp_C")%>% #only select for Month, Day, Year, and temperature variables (drop the day of the week)
  
# We want to replace all the months with their number equivalent (so April is the fourth month of the year, so we replace April with 4)
  mutate(Month = str_replace(Month, "April", "4"))%>%
  mutate(Month = str_replace(Month, "May", "5"))%>%
  mutate(Month = str_replace(Month, "June", "6"))%>%
  mutate(Month = str_replace(Month, "July", "7"))%>%
  mutate(Month = str_replace(Month, "August", "8"))%>%
  mutate(Month = str_replace(Month, "September", "9"))%>%
  mutate(Month = str_replace(Month, "October", "10"))%>%
  mutate(Month = str_replace(Month, "November", "11"))%>%
  mutate(Month = str_replace(Month, "December", "12"))%>%
  drop_na()%>% # This gets rid of some extra rows at the top of the dataset
 
  
# This combines the Month, Day and Year together into one column, separating them with a slash so it looks like 4/1/2021. Then we use the as.Date() function to store the value as a date
  unite("Date", Month:Year, sep = "/")%>%
  mutate(Date = as.Date(Date, "%m/%d/%Y"))%>% 
  filter(Date - start_date >= 0)%>%
  mutate(Temp_C_Avg = as.double(Temp_C_Avg), Temp_C = as.double(Temp_C))%>% #want to store these values as number values not as characters
  
  # We group each of the values by the day so we are able to calculate the average temp for each day
  dplyr::group_by(Date)%>%
  dplyr::summarize(Low = min(Temp_C_Avg), High = max(Temp_C_Avg))


weather
```

```{r}
# Set temperature thresholds and baseline for degree day calculation
baseline = 12
max_high = 29
min_low = 12
threshold = 112

# Function to calculate the average degrees per day
ddm <- function(low, high) {
  average_model <- (low + high)/2 - baseline
  return(average_model)
}

# Create a new column assigning the value as calculated from the average model. If the High is greater than the max_high, 29 will be used to calculate. If the Low is less than the min_low, 12 will be used to calculate
avg_model <- weather %>%
  mutate("Average_Model" = case_when(
    (Low < min_low & High < min_low) ~ ddm(min_low, min_low),
    (Low >= max_high & High >= max_high) ~ ddm(max_high, max_high),
    (Low >= min_low & High <= max_high) ~ ddm(Low, High), 
    (Low < min_low & High <= max_high) ~ ddm(min_low, High),
    (Low >= min_low & High > max_high) ~ ddm(Low, max_high),
    (Low < min_low & High > max_high) ~ ddm(min_low, max_high),
    TRUE ~ ddm(Low, High)
  ))

# Variable to keep track of the cumulative average
cumulative_average = 0

# Empty vector for the generation dates to be added to
generation_dates = c()

# The for loop calculates the cumulative average 
for (num in 1:length(avg_model$Average_Model)) {
  # this takes the average model value calculated above and adds to the cumulative average variable
  cumulative_average = cumulative_average + avg_model$Average_Model[num]
 
  # if the cumulative average exceeds the threshold of 112, it will take that date and add it to the generation_dates vector. it also resets the cumulative average
   if (cumulative_average >= threshold) {
    generation_dates = append(generation_dates, as.Date(avg_model$Date[num]))
    cumulative_average = 0
  }
}

print(generation_dates)

# Create new column with generation information and pesticide information
generation_start_July13 = avg_model %>%
  mutate(generation = case_when(
    Date >= as.Date("2021-07-13") & Date < as.Date("2021-07-21") ~ 1,
    Date >= as.Date("2021-07-21") & Date < as.Date("2021-07-31") ~ 2,
    Date >= as.Date("2021-07-31") & Date < as.Date("2021-08-11") ~ 3,
    Date >= as.Date("2021-08-11") & Date < as.Date("2021-08-20") ~ 4,
    Date >= as.Date("2021-08-20") & Date < as.Date("2021-08-29") ~ 5,
    Date >= as.Date("2021-08-29") & Date < as.Date("2021-09-09") ~ 6,
    Date >= as.Date("2021-09-09") & Date < as.Date("2021-09-19") ~ 7,
    Date >= as.Date("2021-09-19") & Date < as.Date("2021-10-02") ~ 8,
    Date >= as.Date("2021-10-02") & Date < as.Date("2021-10-15") ~ 9,
    Date >= as.Date("2021-10-15") & Date < as.Date("2021-11-11") ~ 10,
    
    TRUE ~ NA_integer_  # Assign NA for dates outside the specified ranges
    ),
      dose_ppm = case_when(
    Date >= as.Date("2021-06-21") & Date <= as.Date("2021-07-20") ~ 0,
    Date >= as.Date("2021-07-21") & Date <= as.Date("2021-08-24") ~ 2.5,
    Date >= as.Date("2021-08-25") & Date <= as.Date("2021-09-06") ~ 7.5,
    Date >= as.Date("2021-09-07") & Date <= as.Date("2021-12-09") ~ 0,
    TRUE ~ NA_real_  # Assign NA for dates outside the specified ranges
    ))


write.csv(generation_start_July13,"./data/generation_start_July13.csv")
```

## References

1.  Information for Degree Day model: <https://treefruit.wsu.edu/crop-protection/opm/dd-models/>
2.  Behrman and Schmidt paper using Degree Day model in Drosophila: <https://www.biorxiv.org/content/10.1101/2022.10.27.514123v1>
